---
title: "Heart Disease Predictor"
author: "Rodrick Mpofu"
date: December 7, 2022
format: 
  revealjs:
    theme: serif
output:
  revealjs::revealjs_presentation:
    transition: fade
    background_transition: slide
---

## Heart Disease Introduction {.smaller background-image="https://pin.it/6p9yshr" background-size="100px" background-repeat="repeat"}

-   918 patients

-   Data from Kaggle collected from five datasets around the world:

    -   Cleveland: 303 observations
    -   Hungarian: 294 observations
    -   Switzerland: 123 observations
    -   Long Beach VA: 200 observations
    -   Stalog (Heart) Data Set: 270 observations

## Variables of interest {.smaller}

-   Variables of interest
    -   `Age`: age of the patient \[years\]

    -   `Sex`: sex of the patient \[M: Male, F: Female\]

    -   `ChestPainType`: chest pain type \[TA: Typical Angina, ATA: Atypical Angina, NAP: Non-Anginal Pain, ASY: Asymptomatic\]

    -   `RestingBP`: resting blood pressure \[mm Hg\]

    -   `Cholesterol`: serum cholesterol \[mm/dl\]

    -   `FastingBS`: fasting blood sugar \[1: if FastingBS \> 120 mg/dl, 0: otherwise\]

    -   `RestingECG`: resting electrocardiogram results \[Normal: Normal, ST: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of \> 0.05 mV), LVH: showing probable or definite left ventricular hypertrophy by Estes' criteria\]

    -   `MaxHR`: maximum heart rate achieved \[Numeric value between 60 and 202\]

    -   `HeartDisease`: output class \[1: heart disease, 0: Normal\]

## Questions that I had {.smaller}

-   Are the cholesterol values higher in males or females?
    -   They are higher in female identifying people.
-   Are females more likely to have Exercise Angina?
    -   Males has more likely to have exercise Angina
-   Are females more likely to have heart-related diseases than males?
    -   Females are less likely than males to have cardiovascular diseases
-   Does the type of chest pain influence the maximum heart rate
    -   Based on the dataset, ATA type of chest pain has more heart rate ( patients with ATA should be put on a medication and special medical attention to reduce their risk of disease)

## Data Exploration {.smaller}

### Age Distribution
```{r}
library(tidyverse)
library(here)
heart_disease_df <- read_csv(here("data/heart_Disease_Dataset_Kaggle.csv"), 
                             col_types = cols(Age = col_double(),
  Sex = col_factor(),
  ChestPainType = col_factor(levels = NULL, ordered = FALSE, include_na = FALSE),
  RestingBP = col_double(),
  Cholesterol = col_double(),
  FastingBS = col_factor(),
  RestingECG = col_factor(),
  MaxHR = col_double(),
  ExerciseAngina = col_factor(),
  Oldpeak = col_double(),
  ST_Slope = col_factor(),
  HeartDisease = col_factor(levels = NULL, ordered = FALSE, include_na = FALSE)))

age_outliers <- boxplot.stats(heart_disease_df$Age)$out

restingBP_outliers <- boxplot.stats(heart_disease_df$RestingBP)$out

choles_outliers <- boxplot.stats(heart_disease_df$Cholesterol)$out

maxHR_outliers <- boxplot.stats(heart_disease_df$MaxHR)$out

oldPeak_outliers <- boxplot.stats(heart_disease_df$Oldpeak)$out


## which function to find which row 
outlier_row <- which(heart_disease_df$RestingBP %in% c(restingBP_outliers) |
                       heart_disease_df$Cholesterol %in% c(choles_outliers) |
                        heart_disease_df$MaxHR %in% c(maxHR_outliers) |
                        heart_disease_df$Oldpeak %in% c(oldPeak_outliers))

heart_disease_df_no_outliers <-heart_disease_df[-outlier_row, ]

## Cholesterol seems to still have outliers

heart_disease_df_no_2 <- 
  heart_disease_df_no_outliers |> filter(Cholesterol > 206 & Cholesterol < 272)

ggplot(heart_disease_df_no_2, aes(x=Age)) +
  geom_histogram(binwidth = 2, colour = "black", fill = "green")
  
```

## Cholesterol vs. Age between sex

```{r}


ggplot(heart_disease_df_no_2, aes(x= Age, y = Cholesterol, color = Sex, shape = Sex)) +
  geom_point() + 
  geom_smooth(color = "purple") +
  scale_colour_brewer(palette = "Accent") +
  facet_wrap(~Sex) 
```

## Proportion of Heart disease between Sex {.smaller}

```{r}

ggplot(data = heart_disease_df_no_2, aes(x = Sex, fill = HeartDisease)) +
  geom_bar(position = "fill") +
  ylab("Proportion")

```

## Proportion of Heart disease between Sex {.smaller}

```{r}
library(GGally)

## pie chart 

pie1<- heart_disease_df_no_2 |> 
  select(ChestPainType, HeartDisease, Sex) |>
  group_by(HeartDisease, Sex) |>
  summarise(n = n()) |>
  mutate(perc = (n/sum(n)) * 100)

ggplot(pie1, aes(x = "", y = perc, fill = Sex)) +
  geom_col() +
  geom_label(aes(label = round(perc, digits=2)),
             position = position_stack(vjust = 0.5),
             show.legend = FALSE,
              )+
  coord_polar(theta = "y") + 
  facet_wrap(~ HeartDisease)

```

## ChestPain type vs MaxHR {.smaller}

```{r}

ggplot(heart_disease_df_no_2, 
       aes(x = ChestPainType, y = MaxHR, color= Sex)) +
  geom_boxplot()

```

## Heatmap

```{r}
# library(tidyverse)
# heat <- heart_disease_df_no_2[ , c(1, 10, 6, 4, 8, 5, 12)] |>
#   mutate(Age = as.factor(Age),
#          Oldpeak = as.factor(Oldpeak),
#          RestingBP = as.factor(RestingBP),
#          MaxHR = as.factor(MaxHR),
#          Cholesterol = as.factor(Cholesterol))
# library(reshape2)
# 
# data_h <- cor(heart_disease_df_no_2[sapply
#                                     (heart_disease_df_no_2,is.numeric)])
# data1<- melt(data_h)
# 
# 
# ggplot(data1, aes(x = Var1,
#                   y = Var2,
#                   fill = value))+
#   geom_tile() + 
#   scale_fill_gradient2(limit = c(-1,1))

library(corrplot)

M<- cor(heart_disease_df_no_2[ , c(1, 10, 4, 8)],  
    use = "everything")

corrplot(M, method="color")

```

## Conclusion

#### Thank you
