---
title: "Project 1"
author: "RodricK Mpofu"
execute:
  echo: true
  warning: false
output: 
  theme: yeti
format: 
  html:
    self-contained: true
---

Statement of Integrity: "All work presented is my own, and I have followed all rules for collaboration. I have not discussed this project with anyone outside of students currently enrolled in this course, and I have listed any such students at the top of my project."

Signed by: Rodrick Mpofu

```{r}
library(tidyverse)
library(readxl)
library(here)

majors_unclean <- read_xlsx(here("data/graduates_2022.xlsx"), sheet = 1, skip = 1)
majors_only <- majors_unclean |> filter(row_number() %in% 1:90 | `...1` == "GRADUATES") |>
  filter(!is.na(`...1`)) |>
  select(1, 2, starts_with("20")) |>
  select(c(1, 3:25))

majors_graduates <- majors_only |>
  pivot_longer(2:24, names_to = "year", values_to = "majors") |>
  mutate(year = parse_number(str_trunc(year, 7))) |>
  rename(major = "...1", 
         n_majors = majors)

majors_df <- majors_graduates |> filter(major != "GRADUATES")
graduates_df <- majors_graduates |> filter(major == "GRADUATES") |>
  select(-major)

slumajors_df <- left_join(majors_df, graduates_df, by = "year") |> 
  rename(n_majors = n_majors.x,
         n_graduates = n_majors.y)
slumajors_df
```

### Part 1

## Question 1

```{r}

slumajors_df <- slumajors_df |> mutate(major_prop = n_majors / n_graduates)

```

## Question 2

```{r}
## create a separate data frame wth Computer Science and Statistics filtered
label_df_colour <- slumajors_df |> filter((major == "Computer Science" | 
                                             major == "Statistics"))

ggplot(slumajors_df, aes(x = year, y = major_prop, group = major)) +
  geom_line(alpha = 0.2) + 
  labs(y = "prop") + 
  scale_colour_viridis_d() +
  geom_line(data = label_df_colour, aes(colour = major), size = 2) + 
  theme_minimal()

```

## Question 3

```{r}

## First create data set for most popular

pop5_df_colour <- slumajors_df |> 
  filter((year == 2022)) |>
  arrange(desc(n_majors)) |> 
  slice(1:5) 

## filter out the 5 to make a line 
pop5_line_df <- slumajors_df |> 
  filter(major == "Business in the Liberal Arts" |
         major == "Economics" |
         major == "Psychology" |
         major == "Government" |
         major == "Perform. & Communication Arts**")

ggplot(slumajors_df, aes(x = year, y = major_prop, group = major)) +
  geom_line(alpha = 0.2) + 
  labs(y = "prop") + 
  scale_colour_viridis_d() +
  geom_line(data = pop5_line_df, aes(colour = major), size = 2) + 
  theme_minimal()

```

## Alternative ways for Question 3

```{r}

## 1st way
pop5_line_df2 <- slumajors_df |> 
  filter (major %in% c("Business in the Liberal Arts", 
                       "Economics", "Psychology", "Government",
                       "Perform. & Communication Arts**"))

## 2nd way
slumajors_df |> filter(major %in% pop5_df_colour$major)

```


## Question 4

Summary: Computer Science and Statics have gained a small proportion of popularity
from 2000 to 2022. They still have a proportion of less than 0.05 of all graduates 
by the end of 2022. However, 4 of the most popular majors, Perfomance and Communication Arts, Economics, Government and Psychology remained very much popular throughout 
2000 to 2022. Additionally, Business in the Liberal Arts gains or increases in popularity
around 2014 until it becomes the most popular major in 2022. 

## Question 5

```{r}
library(knitr)
yr_df<- slumajors_df |> 
  filter(year >= 2018) |> 
  group_by(major) |> 
  summarise(avg_majors = mean(n_majors)) |> 
  arrange(desc(avg_majors)) |> 
  filter(avg_majors > 15) |>
  kable()

yr_df

```

## Question 6

```{r}

## reprex example
toy_df <- tibble(group = c("A", "B", "A", "B"),
                 val = c(2, 5, 1, 9))
toy_df
toy_df |> group_by(group) |>
  summarise(diffs = diff(val))

avg_major_prop1 <- slumajors_df |> 
  filter(year <= 2010 ) |> 
  group_by(major) |>
  summarise(avg_major_prop = mean(major_prop))

avg_major_prop2 <- slumajors_df |> 
  filter(year >= 2010 ) |> 
  group_by(major) |>
  summarise(avg_major_prop = mean(major_prop))

avg_major_df <- bind_rows(avg_major_prop1, avg_major_prop2)

avg_major_df <- avg_major_df |> group_by(major) |> 
  summarise(diffs = diff(avg_major_prop)) |>
  mutate(prop_change = case_when(diffs > 0.01 ~ "positive",
                                 diffs < - 0.01 ~ "negative",
                                 diffs > -0.01 & diffs < 0.01 ~ "little change"))
  

```

## Question 7

```{r}
positive_change_df <- avg_major_df |> 
  filter(prop_change == "positive") |> 
  arrange(desc(diffs)) |> kable()

positive_change_df

negative_change_df <- avg_major_df |> 
  filter(prop_change == "negative") |> 
  arrange(diffs) |> kable()

negative_change_df
```

## Question 8

Summary: Business in the liberal Arts had the highest average graduating majors
with an average of 94.8 graduates per year since 2018. Most majors popularity proportion 
had a little change. Business in the Liberal Arts had the biggest change, a difference in proportion of 0.094. English had the largest negative change in proportion suggesting that more people opted for other majors such as Business in the Liberal Arts.



### Part 2

a. I am interested in the trend of the popularity of mathematics as major from the year 2000 till 2022. Is there a positive trend or negative trend? And also where did this popularity peak and where was it as its lowest.


b. 

```{r}
avg1 <- slumajors_df |> filter(major == "Mathematics") |>group_by(major, year) |> 
  summarise(avg_major_prop = mean(major_prop))

ggplot(avg1, aes(x= year, y = avg_major_prop)) + 
  geom_line(colour = "red") + labs(y = "Average Proportion", title = "Mathemtics Popularity") + 
  geom_smooth(method=lm) + theme_minimal()


```
  
c. From the graph it shows that mathematics did have a positive trend from the year 2000 till 2022. This is shown by the upward sloping blue trend-line. The peak was in 2014 and lowest in 2001. 

